---
title: "gapminder-hw003"
author: "Nicole Mak"
date: "01/10/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**First, let’s load the data and the packages we will use.**

```{r load}
library(gapminder)
library(tidyverse)
library(dplyr)
library(ggplot2)
```



# Homework \#3 assignment **Task Menu**:

Get the maximum and minimum of GDP per capita for all continents. Look
at the spread of GDP per capita within the continents.

Compute a trimmed mean of life expectancy for different years. Or a
weighted mean, weighting by population. Just try something other than
the plain vanilla mean.

How is life expectancy changing over time on different continents?

Report the absolute and/or relative abundance of countries with low life
expectancy over time by continent: Compute some measure of worldwide
life expectancy – you decide – a mean or median or some other quantile
or perhaps your current age. Then determine how many countries on each
continent have a life expectancy less than this benchmark, for each
year.

Find countries with interesting stories. Open-ended and, therefore,
hard. Promising but unsuccessful attempts are encouraged. This will
generate interesting questions to follow up on in class.

## First exercise:

*Get the maximum and minimum of GDP per capita for all continents. Look
at the spread of GDP per capita within the continents.*

```{r max min gdppercap}
gapminder %>% 
 group_by(continent) %>% 
 summarize(min(gdpPercap), max(gdpPercap))

#structuring data by continent
#then requesting that desired observation (minimum and maximum) be summarized in a table, by continent
```

What we observe is that some continents have a larger disparity than
others do! The range for the continents of Africa and Asia is
staggering….

Let’s make a graphs to illustrate this!
```{r distribution graphs}
gapminder %>% 
 group_by(continent) %>%
 ggplot(aes(continent, log(gdpPercap))) +
 geom_point(alpha = 0.05)
 
gapminder %>% 
 group_by(continent) %>%
 ggplot(aes(continent, log(gdpPercap))) +
 geom_boxplot()
```

## Second exercise

*Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.*

```{r weighted mean calculation}
gapminder %>% 
 group_by(continent, year) %>% 
 summarise(mean_lifeExp_weighted = weighted.mean(lifeExp, pop))

#This is a calculation of the weighted mean life expectancy by population for each continent. The means are separated by year so we can observe trends of change in lifeExp over time.
```

```{r lifeExp over time, weighted}
gapminder %>% 
 group_by(continent, year) %>%
 summarise(mean_lifeExp_weighted = weighted.mean(lifeExp, pop)) %>% 
 ggplot(aes(year, mean_lifeExp_weighted))+
  geom_point(aes(colour = continent))

#graphical representation of second exercise
```


## Third exercise

We can use the above information to make some figures which illustrate how life expectancy changes over time.
How is life expectancy changing over time on different continents?



```{r change in lifeexp over time, graph}
gapminder %>%
  group_by(continent, year) %>% 
  summarise(mean_lifeExp_weighted = weighted.mean(lifeExp, pop)) %>% 
  ggplot(aes(year, mean_lifeExp_weighted))+
  geom_line(aes(colour = continent))

#we can even make it a line plot. More still, we can put a regression line to trend overall trend in weighted lifeExp across all continents.

gapminder %>%
  group_by(continent, year) %>% 
  summarise(mean_lifeExp_weighted = weighted.mean(lifeExp, pop)) %>% 
  ggplot(aes(year, mean_lifeExp_weighted))+
  geom_point(aes(colour = continent))+
  geom_smooth()

```

You can see that most continents are following the same trajectory in life expectancy increase starting from the 50s (slopes are visually comparable). Asia is the only continent to rise above and start to catch up to the continents that it was behind of at teh start of the mid 20th century.

## Fourth exercise

*Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.*

```{r median lifeExp as a benchmark}
#Step 1: compute the median life Exp worldwide per year measured.
gapminder %>%
 group_by(year) %>% 
 summarise(median = median(lifeExp))
 
#Step 2: Determine which countries have a life expectancy below the median.
gapminder %>%
 group_by(year) %>% 
 mutate(median = median(lifeExp)) %>% 
 filter(lifeExp<median)


```



Just experimenting with a different approach to tabulate step 2:
```{r tabulate low and high life expectancy countries}
#We can also use the `if_else` function to help us identify which countries are less than the benchmark median for each year. Let's identify those countries with low life expectancy with the word "lower".

gapminder %>%
 group_by(year) %>% 
 mutate(median = median(lifeExp)) %>%
 mutate (benchmark = if_else(lifeExp<median, "lower", "higher"))

```

Anyways, back to determining the relative abundance of countries with low life expectancy by continent....
```{r}
#Step 3: Visually approximate the number of countries with life expectancy below the median. Use aesthetics to set parameters which demonstrate countries not meeting or exceeding the benchmark. Then use facetting to demonstrate the difference between continents.
 
gapminder %>%
 group_by(year) %>% 
 mutate(median = median(lifeExp)) %>%
 ggplot(aes(year, lifeExp)) +
 geom_jitter(aes(colour = (lifeExp < median)), alpha = 0.5)+
  facet_wrap(~ continent)
```

We can really see the disproportionate number of countries with low life expectancy in Africa and Asia compared to Europe and Oceania.

## Fifth exercise

*Find countries with interesting stories. Open-ended and, therefore,
hard. Promising but unsuccessful attempts are encouraged. This will
generate interesting questions to follow up on in class.*

Let's look at the countries in the 5th percentile for low life expectancy in the 21st century...

```{r fifth percentile calculation}

gapminder %>% 
 filter(year > 1999) %>% 
 summarise(ranker = quantile(lifeExp, prob = c(.05)))


gapminder %>% 
 group_by(continent, country) %>% 
 filter(year > 1999) %>% 
 filter(lifeExp<43.80)
```

```{r text graph fifth percentile}
gapminder %>% 
  group_by(continent, country) %>% 
  filter(year>1999) %>% 
  filter(lifeExp<43.8) %>% 
  ggplot(aes(year, lifeExp)) +
  geom_text(aes(label = country), position = "jitter") +
  facet_wrap("year", scales = "free")
```

So we can see that the 5th percentile for lowest life expectancy is comprised solely of African countries and Afghanistan.
Is this something that existed prior to the 21st century? Is there any countries crossing over away fromt he lowest 5% in the world??

```{r}
gapminder %>% 
 filter(year>1950, year<1960) %>% 
 summarise(ranker = quantile(lifeExp, prob = c(.05)))


gapminder %>% 
 group_by(continent, country) %>% 
 filter(year>1950, year<1960) %>% 
 filter(lifeExp<33.6)
```

```{r}

gapminder %>% 
 group_by(continent, country) %>% 
 filter(year>1950, year<1960) %>% 
 filter(lifeExp<33.6) %>% 
 ggplot(aes(year, lifeExp)) + 
 geom_text(aes(label = country), position = "jitter")+
  facet_wrap("year", scales = "free")
```


So, we can see that there has been some change since the 1950s with respect to which countries occupy the lowest 5%.

It is interesting to see Sierra Leone has been at the bottom 5% in 1952 and into the 21st century. Let's see the change in life expectancy over time for Sierra Leone vs. change in life expectancy for a 95% country.

First, let's identify a 95% country to compare Sierra Leone to.

```{r}
gapminder %>% 
 filter(year>1950, year<1960) %>% 
 summarise(ranker = quantile(lifeExp, prob = c(.95)))

gapminder %>%  
 filter(year>1950, year<1960) %>% 
 filter(lifeExp>69.9)
```


```{r}
gapminder %>% 
 filter(year>1999) %>% 
 summarise(ranker = quantile(lifeExp, prob = c(.95)))

gapminder %>% 
  group_by(country) %>% 
 filter(year>1999) %>% 
 filter(lifeExp>80.5)
```


We see that Canada has been in the 95%. Can we compare the change in life expectancy for these two countries?


```{r change in population tabulate}
gapminder %>% 
 group_by(country) %>% 
 mutate(change = lifeExp - lag(lifeExp)) %>% 
  filter(country == "Sierra Leone"| country =="Canada")

 #first, we group by country Then we add a column to the data which calculates the change in population every 5 years for Sierra Leone and CAnada.
```

Let's see if we can make a graph comparing the change in life expectancy over time for these two countries.

```{r Sierra Leone vs Canada}
gapminder %>% 
 group_by(country) %>% 
 mutate(change = lifeExp - lag(lifeExp)) %>% 
  filter(country == "Sierra Leone"| country =="Canada") %>% 
  ggplot(aes(year, change))+
  geom_point(aes(colour = country))+
  geom_line(aes(colour = country))
```

We see that the countries are fairly similar until 1992, when a major event must have occured in Sierra Leone that lead to a disproportionate number of young people dying. Could it be war? famine? natural disaster?

A google search notes that this is around the beginning of a Civil War.


## That's it. Thanks for reading!
