---
title: "gapminder-hw03"
author: "Nicole Mak"
date: "28/09/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**First, let's load the data and the packages we will use.**

```{r}
library(gapminder)
library(tidyverse)
library(dplyr)
library(ggplot2)
```

# Homework #3 assignment **Task Menu**:

Get the maximum and minimum of GDP per capita for all continents.
Look at the spread of GDP per capita within the continents.


Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.

How is life expectancy changing over time on different continents?

Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.

Find countries with interesting stories. Open-ended and, therefore, hard. Promising but unsuccessful attempts are encouraged. This will generate interesting questions to follow up on in class.

## First exercise:

Get the maximum and minimum of GDP per capita for all continents.
Look at the spread of GDP per capita within the continents.

```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarize(min(gdpPercap), max(gdpPercap))

#structuring data by continent
#then requesting that desired observation (minimum and maximum) be summarized in a table, by continent

```

## Second exercise
Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarise(mean_lifeExp_weighted = weighted.mean(lifeExp, pop))

#This is a calculation of the weighted mean life expectancy by population for each continent. The means are separated by year so we can observe trends of change in lifeExp over time.


```

## Third exercise

We can use the above information to make some figures which illustrate how life expectancy changes over time.

How is life expectancy changing over time on different continents?

```{r}

```

